---
title: "Main Visualization Functions in inboggvegan"
author: "Pieter Verschelde"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Main Visualization Functions in inboggvegan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

The `inboggvegan` package provides ggplot2-based visualization functions for 
multivariate analysis objects. While the vegan package offers excellent ordination 
methods, its base R plotting functions can be limiting. This package bridges that 
gap by providing modern, customizable ggplot2 visualizations.

The two main visualization functions are:

1. **`ggbiplot_vegan()`** - Creates biplots for ordination objects
2. **`ggscreeplot()`** - Creates screeplots showing variance explained by components

Both functions work with:
- Vegan objects: rda, cca, capscale
- Standard R PCA objects: prcomp, princomp

## Load Required Packages

```{r libraries}
library(inboggvegan)
library(vegan)
library(ggplot2)
```

# Screeplot Visualizations with ggscreeplot()

Screeplots show the proportion of variance explained by each component in a 
multivariate analysis. This helps determine how many components to retain.

## Example with prcomp

Let's start with a simple PCA analysis using the built-in USArrests dataset:

```{r scree_prcomp}
# Perform PCA
pca <- prcomp(USArrests, scale. = TRUE)

# Three types of screeplots
p1 <- ggscreeplot(pca, type = "pev")  # Proportion of variance
p2 <- ggscreeplot(pca, type = "cev")  # Cumulative variance
p3 <- ggscreeplot(pca, type = "both") # Both on same plot

# Display the plots
p1
p2
p3
```

## Customizing Screeplots

Since the output is a ggplot object, you can customize it further:

```{r scree_custom}
ggscreeplot(pca, type = "both") +
  ggtitle("Screeplot for USArrests PCA") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Example with Vegan Objects

The function also works with vegan ordination objects:

```{r scree_vegan}
data(varespec)
data(varechem)

# RDA analysis
vare.rda <- rda(varespec ~ N + P + K, data = varechem)

# Screeplot showing constrained and unconstrained axes
ggscreeplot(vare.rda, type = "both") +
  ggtitle("Variance Explained in RDA")
```

# Biplot Visualizations with ggbiplot_vegan()

Biplots simultaneously display observations (sites) and variables (species) 
in ordination space, helping visualize relationships between them.

## Simple PCA Biplot

```{r biplot_simple}
# Basic PCA biplot
ggbiplot_vegan(pca)
```

## PCA with Custom Coloring

```{r biplot_pca_color}
# Create a grouping variable based on murder rates
state_data <- data.frame(
  group = ifelse(USArrests$Murder > median(USArrests$Murder), 
                 "High Murder", "Low Murder")
)

# Biplot with colored groups
ggbiplot_vegan(pca,
               site_data = state_data,
               site_mapping = aes(color = group, label = Row.names))
```

## Vegan RDA Biplot

For vegan objects, we can visualize sites, species, and environmental variables:

```{r biplot_rda}
data(dune)
data(dune.env)

# RDA with environmental variables
dune.rda <- rda(dune ~ A1 + Management, data = dune.env)

# Basic biplot with default settings
ggbiplot_vegan(dune.rda)
```

## Customized RDA Biplot

```{r biplot_rda_custom}
# More detailed biplot showing environmental variables
ggbiplot_vegan(dune.rda,
               site_geom = "point",           # Sites as points
               site_data = dune.env,
               site_mapping = aes(color = Management, label = Row.names),
               species_geom = "text",         # Species as text labels
               species_abbrev = TRUE,         # Abbreviate species names
               bp_geom = "arrow",            # Show environmental variables as arrows
               cn_geom = "text") +           # Show factor centroids as text
  theme_minimal() +
  ggtitle("RDA Biplot of Dune Meadow Vegetation")
```

## Core Customization: Colors, Shapes, Sizes, and Labels

The main strength of `ggbiplot_vegan()` is its flexibility in customizing visual 
properties using ggplot2 aesthetics. This is where the package really improves 
over base R plotting functions.

### Customizing Site Points

You can map grouping variables to different aesthetics:

```{r custom_basics}
# Map Management to colors and Use to shapes
ggbiplot_vegan(dune.rda,
               site_data = dune.env,
               site_mapping = aes(color = Management, 
                                 shape = Use,
                                 label = Row.names),
               species_abbrev = TRUE) +
  ggtitle("Sites Colored by Management, Shaped by Use")
```

### Size Mapping for Continuous Variables

Map continuous variables to point size:

```{r custom_size}
# Map A1 (soil thickness) to point size
ggbiplot_vegan(dune.rda,
               site_data = dune.env,
               site_mapping = aes(color = Management, 
                                 size = A1,
                                 label = Row.names),
               species_geom = "text",
               species_abbrev = TRUE) +
  scale_size_continuous(name = "A1 Horizon Thickness") +
  ggtitle("Point Size Shows Soil Thickness")
```

### Custom Colors and Themes

Combine with ggplot2 functions for complete control:

```{r custom_colors}
ggbiplot_vegan(dune.rda,
               site_data = dune.env,
               site_mapping = aes(color = Management, label = Row.names),
               species_abbrev = TRUE) +
  scale_color_manual(values = c("BF" = "darkblue", 
                                 "HF" = "darkgreen", 
                                 "NM" = "orange", 
                                 "SF" = "darkred")) +
  theme_bw() +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank()) +
  ggtitle("Custom Colors and Theme")
```

### Showing vs Hiding Labels

Control which elements show labels:

```{r custom_labels}
# Sites as points (no labels), species as labels
ggbiplot_vegan(dune.rda,
               site_geom = "point",           # No labels on sites
               site_data = dune.env,
               site_mapping = aes(color = Management),
               species_geom = "text",         # Show species labels
               species_abbrev = TRUE) +
  ggtitle("Species Labeled, Sites as Points Only")
```

## Scaling Options

The `scaling` parameter controls how sites and species are displayed:

- **scaling = 1** (distance scaling): Focuses on relationships among sites
- **scaling = 2** (correlation scaling): Focuses on relationships among species; 
  angles between species vectors approximate correlations

```{r scaling}
# Scaling = 1: Distance biplot (focus on sites)
p1 <- ggbiplot_vegan(dune.rda, scaling = 1) +
  ggtitle("Scaling = 1 (Distance)")

# Scaling = 2: Correlation biplot (focus on species)
p2 <- ggbiplot_vegan(dune.rda, scaling = 2) +
  ggtitle("Scaling = 2 (Correlation)")

p1
p2
```

## Adding Confidence Ellipses and Centroids

```{r ellipses}
ggbiplot_vegan(dune.rda,
               site_data = dune.env,
               site_mapping = aes(color = Management, label = Row.names),
               ellipse_geom = "line",                    # Add ellipses
               ellipse_mapping = aes(color = Management),
               ellipse_level = 0.95,                     # 95% confidence
               centroid_geom = "text",                   # Add group centroids as labels
               centroid_mapping = aes(color = Management, label = Management),
               species_abbrev = TRUE) +
  ggtitle("RDA with 95% Confidence Ellipses and Centroids")
```

## CCA Example

Canonical Correspondence Analysis (CCA) works similarly:

```{r cca}
# CCA with environmental variables
dune.cca <- cca(dune ~ A1 + Moisture + Management, data = dune.env)

ggbiplot_vegan(dune.cca,
               site_data = dune.env,
               site_mapping = aes(color = Management, label = Row.names),
               species_geom = "text",
               species_abbrev = TRUE,
               bp_geom = "arrow") +
  theme_minimal() +
  ggtitle("CCA of Dune Meadow Vegetation")
```

## Correlation Circle

When using scaling = 2, you can add a correlation circle. Species with 
endpoints on or near this circle are well represented by the displayed axes 
(have high communality for these axes):

```{r cor_circle}
ggbiplot_vegan(dune.rda,
               scaling = 2,
               species_abbrev = TRUE,
               cor_circle_geom = "line") +  # Add correlation circle
  ggtitle("RDA with Correlation Circle")
```

## Choosing Different Axes

By default, axes 1 and 2 are displayed. If the ordination object has at least 
three axes, you can plot different axis combinations:

```{r different_axes}
# If there are enough axes, you can plot different combinations
# For example, axes 1 and 3
ggbiplot_vegan(dune.rda, 
               choices = c(1, 3),
               species_abbrev = TRUE) +
  ggtitle("RDA Axes 1 and 3")
```

## Comparison: princomp vs prcomp

Both princomp and prcomp perform PCA, but use different algorithms. 
The package now ensures consistent scaling between them:

```{r princomp_vs_prcomp}
# Using princomp
pca_princomp <- princomp(USArrests, cor = TRUE)
p1 <- ggbiplot_vegan(pca_princomp) + ggtitle("princomp")

# Using prcomp
pca_prcomp <- prcomp(USArrests, scale. = TRUE)
p2 <- ggbiplot_vegan(pca_prcomp) + ggtitle("prcomp")

p1
p2
```

The visualizations should now show consistent scaling, making comparisons easier.

# Summary

The `inboggvegan` package provides flexible ggplot2-based visualizations for 
ordination analysis. Key features include:

- **Easy customization**: Map grouping variables to colors, shapes, and sizes
- **Multiple object types**: Works with prcomp, princomp, rda, cca, capscale
- **Consistent scaling**: Fixed scaling between princomp and prcomp
- **Full ggplot2 compatibility**: Add any ggplot2 layers for complete control

For more details on function parameters, see the function documentation 
(`?ggbiplot_vegan`, `?ggscreeplot`).

# Helper Functions

The package also exports several helper functions for building custom plots:

- `get_display_data()` - Extract and prepare data for plotting
- `geom_arrow()` - Draw arrows for biplot vectors
- `geom_arrowlabel()` - Add labels to arrows
- `geom_corcircle()` - Add correlation circle
- `stat_centroid()` - Calculate and display group centroids
- `geom_selector()` - Flexible geom selection

These can be used to create completely custom visualizations if needed.

# Summary

The `inboggvegan` package provides two main visualization functions:

1. **`ggscreeplot()`** - Visualize variance explained by components
   - Works with prcomp, princomp, rda, cca, capscale (and lda from MASS package)
   - Three types: "pev", "cev", "both"
   - Fully customizable with ggplot2

2. **`ggbiplot_vegan()`** - Create biplots for ordination objects
   - Works with all vegan ordination objects and PCA objects
   - Highly customizable display of sites, species, and environmental variables
   - Multiple geometry options (points, text, arrows)
   - Support for ellipses, centroids, and correlation circles
   - Consistent scaling between princomp and prcomp

Both functions return ggplot objects that can be further customized using 
standard ggplot2 functions, providing maximum flexibility for publication-quality 
visualizations.

# Session Info

```{r session}
sessionInfo()
```
